---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastsimplexreg

<!-- badges: start -->
<!-- badges: end -->

`fastsimplexreg` provides high-performance maximum-likelihood estimation of
**simplex regression** models for continuous proportions in the open interval
$(0, 1)$, following Barndorff-Nielsen and Jorgensen (1991). It fits separate
submodels for the mean and the dispersion,

$$
Y_i \sim \operatorname{Simplex}(\mu_i, \phi_i), \qquad
g(\mu_i) = \mathbf{x}_i^\top \boldsymbol\beta, \qquad
\log(\phi_i) = \mathbf{z}_i^\top \boldsymbol\gamma,
$$

with the density

$$
f(y; \mu, \phi) = \left[2\pi\phi\,(y(1-y))^3\right]^{-1/2}
\exp\!\left\{-\frac{1}{2\phi}\,
\frac{(y-\mu)^2}{y(1-y)\,\mu^2(1-\mu)^2}\right\}, \qquad 0 < y < 1.
$$

The entire numerical hot path -- log-likelihood, analytic score, a native
BFGS optimiser, density, random generation, prediction and link inverses -- is
implemented in C++ with `RcppArmadillo`, BLAS/LAPACK and optional OpenMP
parallelism, so that models scale to large data sets.

## Installation

You can install the development version from
[GitHub](https://github.com/evandeilton/fastsimplexreg):

```r
# install.packages("remotes")
remotes::install_github("evandeilton/fastsimplexreg")
```

## The multi-part formula interface

The API uses the `Formula` package and separates the two submodels with the
`|` operator:

```r
fit <- fastsimplexreg(y ~ x1 + x2 | z1 + z2, data = dat, link = "logit")
```

The first right-hand side component models the mean $\mu$; the second models the
dispersion $\phi$. When the second component is omitted, as in `y ~ x1 + x2`,
the dispersion is constant (equivalent to `| 1`).

## Mean links

The mean supports four links; the dispersion always uses a log link.

| Link      | $g(\mu)$                  | $g^{-1}(\eta)$              |
|-----------|---------------------------|----------------------------|
| `logit`   | $\log\{\mu/(1-\mu)\}$     | $1/(1 + e^{-\eta})$        |
| `probit`  | $\Phi^{-1}(\mu)$          | $\Phi(\eta)$               |
| `cloglog` | $\log\{-\log(1-\mu)\}$    | $1 - \exp(-\exp(\eta))$    |
| `neglog`  | $-\log\{-\log(\mu)\}$     | $\exp(-\exp(-\eta))$       |

The `neglog` definition matches the CRAN `simplexreg` package.

## A worked example

```{r example, eval = requireNamespace("fastsimplexreg", quietly = TRUE)}
library(fastsimplexreg)

set.seed(20260710)
n <- 2000
dat <- data.frame(
  x1 = rnorm(n),
  x2 = rbinom(n, size = 1, prob = 0.35),
  z1 = rnorm(n)
)

mu_true  <- simplex_linkinv(-0.30 + 0.90 * dat$x1 - 0.55 * dat$x2, link = "logit")
phi_true <- exp(-1.10 + 0.65 * dat$z1)
dat$y <- rsimplex(n, mu = mu_true, phi = phi_true)

fit <- fastsimplexreg(y ~ x1 + x2 | z1, data = dat, link = "logit", n_threads = 1L)
summary(fit)
```

Coefficients, the variance-covariance matrix, the log-likelihood and predictions
are available through the usual extractor methods:

```{r extractors, eval = requireNamespace("fastsimplexreg", quietly = TRUE)}
coef(fit)
confint(fit)
logLik(fit)
head(predict(fit, type = "both"))
```

## Prediction

The available prediction types are `"response"`/`"mean"`, `"dispersion"`,
`"link"` and `"both"`. Prediction on new data reuses the `terms`, `xlevels` and
`contrasts` stored at fitting time:

```r
predict(fit, newdata = new_dat, type = "both")
```

## Performance notes

* **Analytic score** for every link, avoiding numerical differentiation during
  optimisation.
* **Native BFGS in C++**, so the optimiser does not cross the R/C++ boundary at
  each objective evaluation.
* **Armadillo/BLAS** linear algebra for the linear predictors.
* **OpenMP** parallelism over observations; `n_threads = 0` requests all
  available threads.
* **A single `model.frame`** for both submodels, guaranteeing consistent
  handling of `subset`, `NA`, factors and levels.
* **Optional inference**: pass `inference = FALSE` to skip the Hessian for
  exploratory fits on massive data.

## References

Barndorff-Nielsen, O. E. and Jorgensen, B. (1991). Some parametric models on
the simplex. *Journal of Multivariate Analysis*, **39**(1), 106--116.

Zhang, P., Qiu, Z. and Shi, C. (2016). simplexreg: An R Package for Regression
Analysis of Proportional Data Using the Simplex Distribution. *Journal of
Statistical Software*, **71**(11), 1--21.

## License

GPL-3.
