---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastsimplexreg

<!-- badges: start -->

<!-- [![CRAN -->
<!-- status](https://www.r-pkg.org/badges/version/fastsimplexreg)](https://CRAN.R-project.org/package=fastsimplexreg) -->
[![R-CMD-check](https://github.com/evandeilton/fastsimplexreg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/evandeilton/fastsimplexreg/actions/workflows/R-CMD-check.yaml)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/fastsimplexreg)](https://cran.r-project.org/package=fastsimplexreg)
[![License:MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

<!-- badges: end -->

A high-performance implementation of simplex regression for continuous
proportions in the open interval $(0, 1)$, with separate submodels for the mean
and the dispersion, and both fixed- and mixed-effects fitting. The numerical
core (log-likelihood, analytic score, native BFGS, adaptive Gauss-Hermite
quadrature, density, random generation, prediction) is written in C++ with
`RcppArmadillo`, `BLAS`/`LAPACK` and optional `OpenMP`.

`fastsimplexreg` provides high-performance maximum-likelihood estimation of
**simplex regression** models for continuous proportions in the open interval
$(0, 1)$, following Barndorff-Nielsen and Jorgensen (1991). It fits separate
submodels for the mean and the dispersion,

$$
Y_i \sim \mathrm{Simplex}(\mu_i, \phi_i), \qquad
g(\mu_i) = \mathbf{x}_i^\top \boldsymbol\beta, \qquad
\log(\phi_i) = \mathbf{z}_i^\top \boldsymbol\gamma,
$$

with the density

$$
f(y; \mu, \phi) = \left[2\pi\phi\,(y(1-y))^3\right]^{-1/2}
\exp\!\left\lbrace -\frac{1}{2\phi}\,
\frac{(y-\mu)^2}{y(1-y)\,\mu^2(1-\mu)^2}\right\rbrace, \qquad 0 < y < 1.
$$

For nested or clustered proportion data, the package also fits **simplex mixed-effects models** with cluster-specific random effects $\mathbf{b}_j \sim N_q(\mathbf{0}, \boldsymbol\Sigma)$ in the mean submodel:

$$
Y_{ij} \mid \mathbf{b}_j \sim \mathrm{Simplex}(\mu_{ij}, \phi_{ij}), \qquad
g(\mu_{ij}) = \mathbf{x}_{ij}^\top \boldsymbol\beta + \mathbf{z}_{ij}^\top \mathbf{b}_j, \qquad
\log(\phi_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol\gamma.
$$

Since the random effects are unobserved, estimation is performed by maximizing the marginal likelihood:

$$
L(\boldsymbol\beta, \boldsymbol\gamma, \boldsymbol\Sigma) = \prod_j \int_{\mathbb{R}^q} \left[ \prod_i f(y_{ij} \mid \mathbf{b}_j; \mu_{ij}, \phi_{ij}) \right] f(\mathbf{b}_j; \boldsymbol\Sigma) \, \mathrm{d}\mathbf{b}_j,
$$

where $f(\mathbf{b}_j; \boldsymbol\Sigma)$ is the multivariate normal density. This high-dimensional integral is approximated using high-performance **adaptive Gauss-Hermite quadrature** (AGHQ).

The entire numerical hot path -- log-likelihood, analytic score, a native
BFGS optimiser, density, random generation, prediction and link inverses -- is
implemented in C++ with `RcppArmadillo`, BLAS/LAPACK and optional OpenMP
parallelism, so that models scale to large data sets.

## Installation

You can install the development version from
[GitHub](https://github.com/evandeilton/fastsimplexreg):

```r
# install.packages("remotes")
remotes::install_github("evandeilton/fastsimplexreg")
```

## The multi-part formula interface

The API uses the `Formula` package and separates the two submodels with the
`|` operator:

```r
fit <- fastsimplexreg(y ~ x1 + x2 | z1 + z2, data = dat, link = "logit")
```

The first right-hand side component models the mean $\mu$; the second models the
dispersion $\phi$. When the second component is omitted, as in `y ~ x1 + x2`,
the dispersion is constant (equivalent to `| 1`).

## Mean links

The mean supports four links; the dispersion always uses a log link.

| Link      | $g(\mu)$                  | $g^{-1}(\eta)$              |
|-----------|---------------------------|----------------------------|
| `logit`   | $\log\{\mu/(1-\mu)\}$     | $1/(1 + e^{-\eta})$        |
| `probit`  | $\Phi^{-1}(\mu)$          | $\Phi(\eta)$               |
| `cloglog` | $\log\{-\log(1-\mu)\}$    | $1 - \exp(-\exp(\eta))$    |
| `neglog`  | $-\log\{-\log(\mu)\}$     | $\exp(-\exp(-\eta))$       |

The `neglog` definition follows Zhang et al. (2016).

## A worked example

We use the `ReadingSkills` data from the **betareg** package: the response
`accuracy` is a reading-accuracy score in $(0, 1)$ for children with and without
`dyslexia`, together with a standardised `iq`. We let the mean depend on both and
the dispersion depend on dyslexia status.

```{r example, eval = requireNamespace("betareg", quietly = TRUE)}
library(fastsimplexreg)
data("ReadingSkills", package = "betareg")

fit <- fastsimplexreg(accuracy ~ dyslexia + iq | dyslexia,
                      data = ReadingSkills, link = "logit")
summary(fit)
```

Coefficients, the variance-covariance matrix, the log-likelihood and predictions
are available through the usual extractor methods:

```{r extractors, eval = requireNamespace("betareg", quietly = TRUE)}
coef(fit)
confint(fit)
logLik(fit)
head(predict(fit, type = "both"))
```

## Mixed-effects models

For nested or clustered proportion data, `fastsimplexregmixed()` adds a
cluster-specific random effect, estimated by adaptive Gauss-Hermite quadrature.
The `GasolineYield` data (also from **betareg**) records the proportion `yield`
of crude oil converted to gasoline across several experimental conditions, in 10
crude-oil `batch`es. We fit a random intercept per batch:

```{r mixed, eval = requireNamespace("betareg", quietly = TRUE)}
data("GasolineYield", package = "betareg")

mfit <- fastsimplexregmixed(yield ~ temp, random = ~ 1 | batch,
                            data = GasolineYield, link = "logit", nAGQ = 15)
summary(mfit)
VarCorr(mfit)
```

## Prediction

The available prediction types are `"response"`/`"mean"`, `"dispersion"`,
`"link"` and `"both"`. Prediction on new data reuses the `terms`, `xlevels` and
`contrasts` stored at fitting time:

```r
predict(fit, newdata = new_dat, type = "both")
```

## Performance notes

* **Analytic score** for every link, avoiding numerical differentiation during
  optimisation.
* **Native BFGS in C++**, so the optimiser does not cross the R/C++ boundary at
  each objective evaluation.
* **Armadillo/BLAS** linear algebra for the linear predictors.
* **OpenMP** parallelism over observations; `n_threads = 0` requests all
  available threads.
* **A single `model.frame`** for both submodels, guaranteeing consistent
  handling of `subset`, `NA`, factors and levels.
* **Optional inference**: pass `inference = FALSE` to skip the Hessian for
  exploratory fits on massive data.

In head-to-head benchmarks, `fastsimplexreg` returns **numerically identical
maximum-likelihood estimates** to the CRAN package `simplexreg` (fitted-mean
correlation = 1) while running roughly **5-13x faster** for sample sizes up to
1e5, and is ~2-4x faster than `betareg` while fitting a richer model. See the
[benchmark article](https://evandeilton.github.io/fastsimplexreg/articles/benchmark.html)
(the full reproducible study ships in `inst/benchmark/`).

## References

Barndorff-Nielsen, O. E. and Jorgensen, B. (1991). Some parametric models on
the simplex. *Journal of Multivariate Analysis*, **39**(1), 106--116.

Zhang, P., Qiu, Z. and Shi, C. (2016). simplexreg: An R Package for Regression
Analysis of Proportional Data Using the Simplex Distribution. *Journal of
Statistical Software*, **71**(11), 1--21.

## License

MIT © José Evandeilton Lopes.
