---
title: "Simplex mixed models for nested proportion data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simplex mixed models for nested proportion data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fastsimplexreg)
```

## 1. Motivation

Proportion responses in the open interval $(0, 1)$ are frequently **nested**:
repeated measurements within subjects, students within schools, plots within
sites. Ignoring the clustering understates uncertainty and can bias the mean and
dispersion estimates. `fastsimplexregmixed()` fits a two-level simplex mixed
model with a random effect per cluster while retaining the variable-dispersion
structure of [`fastsimplexreg()`](fastsimplexreg.html).

## 2. The model

For observation $i$ in cluster $j$, conditional on a cluster random effect
$b_j \sim N_q(0, \Sigma)$,
$$
y_{ij} \mid b_j \sim \mathrm{Simplex}(\mu_{ij}, \phi_{ij}), \qquad
g(\mu_{ij}) = x_{ij}^\top\beta + z_{ij}^\top b_j, \qquad
\log\phi_{ij} = w_{ij}^\top\gamma .
$$
The mean link $g$ is one of `logit`, `probit`, `cloglog`, `neglog`; the
dispersion link is log. The random effects enter the mean submodel; the
dispersion submodel carries fixed effects only.

The marginal likelihood integrates the random effects out,
$$
\ell(\theta) = \sum_j \log \int_{\mathbb{R}^q}
  \exp\Big\{\textstyle\sum_i \log f(y_{ij}\mid b)\Big\}\, N(b; 0, \Sigma)\, \mathrm{d}b ,
$$
which has no closed form and is approximated by **adaptive Gauss-Hermite
quadrature** (AGHQ). With `nAGQ = 1` this is the Laplace approximation; larger
`nAGQ` refines the integral. The covariance $\Sigma = D D^\top$ is estimated on
an unconstrained log-Cholesky scale, so the estimate is always positive
definite.

## 3. The `random` interface

Random effects and the grouping factor are given through `random`, using the
familiar lme4-style bar:

* `random = ~ 1 | subject` — a random intercept;
* `random = ~ 1 + x1 | subject` — a random intercept and a random slope in `x1`;
* `random = ~ 0 + x1 | subject` — a random slope only.

The mean and dispersion fixed effects use the same multi-part `Formula` as
`fastsimplexreg()`, e.g. `y ~ x1 + x2 | z1`.

## 4. A worked example

```{r simulate}
set.seed(2026)
J <- 150; nj <- 10; n <- J * nj
dat <- data.frame(
  subject = factor(rep(seq_len(J), each = nj)),
  x1 = rnorm(n),
  z1 = rnorm(n)
)
b <- rnorm(J, 0, 0.7)[dat$subject]          # true random intercepts
mu  <- simplex_linkinv(0.4 - 0.7 * dat$x1 + b, "logit")
phi <- exp(-0.3 + 0.4 * dat$z1)
dat$y <- rsimplex(n, mu, phi)
```

```{r fit}
fit <- fastsimplexregmixed(
  y ~ x1 | z1,
  random = ~ 1 | subject,
  data = dat,
  link = "logit",
  nAGQ = 9,
  n_threads = 1
)
summary(fit)
```

The estimated variance component and the group-level random effects:

```{r ranef}
VarCorr(fit)
head(ranef(fit))
```

Predictions are conditional on the estimated random effects by default; set
`re.form = NA` for population-level (marginal-mode) predictions:

```{r predict}
head(predict(fit, type = "response"))                 # includes random effects
head(predict(fit, type = "response", re.form = NA))   # population level
```

Diagnostic plots (built with **ggplot2**, residuals conditional on the random
effects):

```{r plot, fig.width = 7, fig.height = 5.5, out.width = "100%"}
plot(fit, which = 1:4)
```

## 5. Choosing `nAGQ` and performance

`nAGQ` trades accuracy for cost: `nAGQ = 1` (Laplace) is the fastest and is often
adequate when clusters are large, while `nAGQ = 9`-`15` gives high accuracy for
small clusters. The tensor grid has `nAGQ^q` nodes, so keep `q` small (`q <= 3`)
and reduce `nAGQ` as `q` grows (`11` for `q = 1`, `9` for `q = 2`, `7` for
`q = 3` are good defaults). Because clusters are conditionally independent, the
work is parallelised over clusters; set `n_threads = 0` to use all available
cores.

## 6. Scope and extensions

This version supports a single grouping factor (two-level nesting), Gaussian
random effects in the mean, and fixed-effect dispersion. Crossed or three-level
random effects, and random effects in the dispersion submodel, are natural
extensions for future releases.

## References

Barndorff-Nielsen, O. E. and Jørgensen, B. (1991). Some parametric models on the
simplex. *Journal of Multivariate Analysis*, **39**(1), 106-116.

Pinheiro, J. C. and Bates, D. M. (1995). Approximations to the log-likelihood
function in the nonlinear mixed-effects model. *Journal of Computational and
Graphical Statistics*, **4**(1), 12-35.
