---
title: "Fast simplex regression with variable dispersion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast simplex regression with variable dispersion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(fastsimplexreg)
```

## 1. The simplex distribution

`fastsimplexreg` implements maximum-likelihood regression for continuous
responses on the open unit interval $(0, 1)$ using the **simplex distribution**
of Barndorff-Nielsen and Jørgensen (1991). A response $Y$ with mean
$\mu \in (0, 1)$ and dispersion $\phi > 0$ has density

$$
f(y; \mu, \phi) =
\bigl[\, 2\pi\,\phi\,\{y(1-y)\}^{3} \,\bigr]^{-1/2}
\exp\!\left\{
-\frac{1}{2\phi}\,
\frac{(y-\mu)^2}{y(1-y)\,\mu^2(1-\mu)^2}
\right\},
\qquad 0 < y < 1 .
$$

The parameter $\mu$ is the mean of $Y$, while $\phi$ (equivalently
$\sigma^2$) controls the dispersion. It is convenient to work with the
**log-density**

$$
\log f(y; \mu, \phi) =
-\tfrac{1}{2}\bigl(\log 2\pi + \log \phi\bigr)
-\tfrac{3}{2}\bigl(\log y + \log(1-y)\bigr)
-\frac{1}{2\phi}\, d(y, \mu),
\qquad
d(y, \mu) = \frac{(y-\mu)^2}{y(1-y)\,\mu^2(1-\mu)^2}.
$$

The density and its logarithm are available through `dsimplex()`, and random
variates through `rsimplex()`:

```{r density}
# Density on a grid and a quick check that it integrates to one.
x <- seq(0.01, 0.99, length.out = 99)
d <- dsimplex(x, mu = 0.4, phi = 1, n_threads = 1L)
integrate(function(t) dsimplex(t, mu = 0.4, phi = 1), 0, 1)$value
```

```{r rng}
set.seed(1)
y <- rsimplex(2000L, mu = 0.4, phi = 1)
c(sample_mean = mean(y), target_mu = 0.4)
```

## 2. The regression model

Given covariates, each observation $Y_i$ follows a simplex distribution with its
own mean $\mu_i$ and dispersion $\phi_i$. Two linear submodels are fitted
jointly:

$$
g(\mu_i) = \mathbf{x}_i^\top \boldsymbol\beta,
\qquad
\log(\phi_i) = \mathbf{z}_i^\top \boldsymbol\gamma .
$$

The mean submodel uses one of four link functions $g$, and the dispersion
submodel always uses the logarithmic link, guaranteeing $\phi_i > 0$.

| Link | $g(\mu)$ | $g^{-1}(\eta)$ |
|------|----------|----------------|
| `logit`   | $\log\{\mu/(1-\mu)\}$ | $1/(1 + e^{-\eta})$ |
| `probit`  | $\Phi^{-1}(\mu)$ | $\Phi(\eta)$ |
| `cloglog` | $\log\{-\log(1-\mu)\}$ | $1 - \exp\{-\exp(\eta)\}$ |
| `neglog`  | $-\log\{-\log(\mu)\}$ | $\exp\{-\exp(-\eta)\}$ |

The `neglog` link follows the definition used by the CRAN package
`simplexreg`. All four inverse links are exposed through `simplex_linkinv()`:

```{r links}
eta <- c(-1, 0, 1)
sapply(c("logit", "probit", "cloglog", "neglog"),
       function(lk) simplex_linkinv(eta, lk))
```

## 3. The multi-part formula API

The mean and dispersion submodels are specified in a single multi-part formula
from the `Formula` package, with the two parts separated by the `|` operator:

```r
y ~ x1 + x2 | z1 + z2
```

The first right-hand-side part models the mean $\mu$; the second models the
dispersion $\phi$. Omitting the second part, as in `y ~ x1 + x2`, assumes a
constant dispersion (equivalent to `| 1`). Both components are built from a
single shared `model.frame`, so `subset`, missing-value handling, factors and
their levels are treated consistently across $X$ and $Z$, avoiding any
misalignment between the two design matrices.

## 4. A worked example

We simulate a moderate data set with a known mean structure (logit link) and a
covariate-dependent dispersion (log link), fit the model, and inspect the
results. Everything runs on a single thread for reproducibility.

```{r simulate}
set.seed(20260710)
n <- 2000L
dat <- data.frame(
  x1 = rnorm(n),
  x2 = rbinom(n, size = 1L, prob = 0.4),
  z1 = rnorm(n)
)

eta_mu   <- -0.3 + 0.9 * dat$x1 - 0.5 * dat$x2
mu_true  <- simplex_linkinv(eta_mu, link = "logit")
phi_true <- exp(-0.8 + 0.6 * dat$z1)
dat$y    <- rsimplex(n, mu = mu_true, phi = phi_true)
```

```{r fit}
fit <- fastsimplexreg(
  y ~ x1 + x2 | z1,
  data = dat,
  link = "logit",
  n_threads = 1L
)
fit
```

The estimated coefficients recover the data-generating values. A full inference
table, including standard errors and Wald tests, is produced by `summary()`:

```{r summary}
summary(fit)
```

Standard extractors work as expected:

```{r extractors}
coef(fit)
logLik(fit)
c(AIC = AIC(fit), BIC = BIC(fit), nobs = nobs(fit))
```

Wald confidence intervals are available through `confint()`:

```{r confint}
confint(fit, level = 0.95)
```

Prediction supports several output types. With `newdata` the stored `terms`,
`xlevels`, and `contrasts` are reused, so predictions on fresh data are
consistent with the fitted model:

```{r predict}
head(predict(fit, type = "both"))
new <- data.frame(x1 = c(-1, 0, 1), x2 = c(0, 1, 0), z1 = c(0, 0, 0))
predict(fit, newdata = new, type = "response")
```

Residuals are available on the response and Pearson scales:

```{r residuals}
summary(residuals(fit, type = "pearson"))
```

## 5. Performance and design notes

`fastsimplexreg` is designed for large data sets and keeps the entire critical
path in C++:

- **C++ log-likelihood and analytic score.** The derivative of the
  log-likelihood with respect to $\mu$ is computed in closed form and combined
  with the link-specific derivative $\partial\mu/\partial\eta_\mu$ via the chain
  rule,
  $$
  \frac{\partial \ell_i}{\partial \eta_{\mu_i}}
  = \frac{\partial \ell_i}{\partial \mu_i}\,
    \frac{\partial \mu_i}{\partial \eta_{\mu_i}} ,
  $$
  which avoids numerical differentiation during optimisation.
- **Native BFGS.** The optimiser iterates entirely in C++, avoiding repeated
  crossings of the R/C++ boundary at every objective evaluation.
- **BLAS matrix-vector products.** The linear predictors
  $\boldsymbol\eta_\mu = X\boldsymbol\beta$ and
  $\boldsymbol\eta_\phi = Z\boldsymbol\gamma$ use Armadillo/BLAS.
- **Optional OpenMP.** The per-observation likelihood and score evaluation can
  be parallelised. Passing `n_threads = 0L` requests the maximum number of
  threads available to the backend; random number generation always runs in
  serial to preserve reproducibility.
- **Optional inference.** The Hessian, covariance matrix, and standard errors
  are computed only when `inference = TRUE` (the default). For exploratory fits
  on massive data, set `inference = FALSE` to skip this step.

```{r perf-note, eval = FALSE}
# On large data, request all available threads and skip inference for speed:
fit_fast <- fastsimplexreg(
  y ~ x1 + x2 | z1,
  data = dat,
  link = "logit",
  n_threads = 0L,
  inference = FALSE
)
```

## Model diagnostics

The `plot()` method draws diagnostic panels built with **ggplot2**: residuals
against the fitted mean, a normal Q-Q plot, a scale-location plot and observed
against fitted values. By default it uses deviance residuals; the panels are
combined with **patchwork** when that package is installed.

```{r diagnostics, fig.width = 7, fig.height = 5.5, out.width = "100%"}
plot(fit, which = 1:4)
```

Several standard extractor methods are available, mirroring `lm`/`glm` fits:

```{r methods}
deviance(fit)
head(residuals(fit, type = "deviance"))
AIC(fit)
BIC(fit)
head(model.matrix(fit, model = "mean"))
# simulate new responses from the fitted model
sims <- simulate(fit, nsim = 2, seed = 1)
head(sims)
# refit through an updated (multi-part) formula
update(fit, . ~ . - x2 | z1)
```

## References

Barndorff-Nielsen, O. E. and Jørgensen, B. (1991). Some parametric models on
the simplex. *Journal of Multivariate Analysis*, **39**(1), 106–116.
