---
title: "Fast simplex regression with variable dispersion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast simplex regression with variable dispersion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(fastsimplexreg)
has_betareg <- requireNamespace("betareg", quietly = TRUE)
```

## 1. Bounded proportions and the simplex distribution

Responses that are continuous **proportions** or **rates** restricted to the open
interval $(0, 1)$ — the fraction of a resource used, a recovery rate, a
concentration relative to a maximum — violate the assumptions of ordinary linear
regression: their support is bounded and their variance is inherently
heteroscedastic, shrinking towards the boundaries $0$ and $1$. Two likelihood
based models dominate this setting: **beta regression** and **simplex
regression**. This package implements the latter.

The simplex distribution (Barndorff-Nielsen and Jørgensen, 1991) is a member of the class of *dispersion
models*. A random variable $Y \in (0, 1)$ has a simplex distribution with mean
$\mu \in (0, 1)$ and dispersion $\phi > 0$, written
$Y \sim \mathrm{Simplex}(\mu, \phi)$, if its density is
$$
f(y; \mu, \phi) = \bigl[2\pi\,\phi\,\{y(1-y)\}^3\bigr]^{-1/2}
  \exp\!\left\{ -\frac{1}{2\phi}\, d(y; \mu) \right\},
\qquad
d(y; \mu) = \frac{(y-\mu)^2}{y(1-y)\,\mu^2(1-\mu)^2},
$$
where $d(y; \mu)$ is the *unit deviance*. On the log scale,
$$
\log f(y; \mu, \phi) = -\tfrac12\bigl(\log 2\pi + \log\phi\bigr)
  - \tfrac32\bigl(\log y + \log(1-y)\bigr) - \tfrac{1}{2\phi}\, d(y; \mu).
$$
The mean of $Y$ is exactly $\mu$, and to first order in the dispersion the
variance is $\mathrm{Var}(Y) \approx \phi\, V(\mu)$ with the simplex **variance
function** $V(\mu) = \mu^3(1-\mu)^3$ — the variance is largest near
$\mu = 1/2$ and vanishes at the boundaries. The dispersion $\phi$ here is the
$\sigma^2$ parameter of Barndorff-Nielsen and Jørgensen (1991).

```{r density-shape, fig.width = 6, fig.height = 3.4, out.width = "85%"}
grid <- seq(0.001, 0.999, length.out = 400)
op <- par(mar = c(4, 4, 1, 1))
plot(grid, dsimplex(grid, mu = 0.5, phi = 1), type = "l", lwd = 2,
     xlab = "y", ylab = "density", ylim = c(0, 4))
lines(grid, dsimplex(grid, mu = 0.5, phi = 0.3), lwd = 2, lty = 2, col = "#1a5276")
lines(grid, dsimplex(grid, mu = 0.3, phi = 1),   lwd = 2, lty = 3, col = "#a93226")
legend("topright", bty = "n",
       legend = c("mu=0.5, phi=1", "mu=0.5, phi=0.3", "mu=0.3, phi=1"),
       lwd = 2, lty = 1:3, col = c("black", "#1a5276", "#a93226"))
par(op)
```

Smaller $\phi$ concentrates the mass around the mean; changing $\mu$ shifts and
skews the density.

## 2. The regression model

`fastsimplexreg()` models both the mean and the dispersion as functions of
covariates. For observation $i$,
$$
Y_i \sim \mathrm{Simplex}(\mu_i, \phi_i), \qquad
g(\mu_i) = x_i^\top\beta, \qquad
\log\phi_i = z_i^\top\gamma.
$$
The dispersion always uses a log link (so $\phi_i > 0$). The mean link $g$ can be
**logit**, **probit**, **cloglog** or **neglog**:

| link | $g(\mu)$ | $g^{-1}(\eta)$ |
|------|----------|----------------|
| logit | $\log\{\mu/(1-\mu)\}$ | $1/(1+e^{-\eta})$ |
| probit | $\Phi^{-1}(\mu)$ | $\Phi(\eta)$ |
| cloglog | $\log\{-\log(1-\mu)\}$ | $1 - \exp(-e^{\eta})$ |
| neglog | $-\log\{-\log\mu\}$ | $\exp(-e^{-\eta})$ |

Estimation is by maximum likelihood. The log-likelihood, the **analytic score**,
a native **BFGS** optimiser, and the link inverses all run in C++; the
per-observation loop can use OpenMP, so the fit scales to very large data sets.

### The multi-part formula interface

The mean and dispersion submodels are separated by `|` in a
[`Formula`](https://CRAN.R-project.org/package=Formula):

```r
fastsimplexreg(y ~ x1 + x2 | z1 + z2, data = dat, link = "logit")
```

Here `y ~ x1 + x2` is the mean model and `z1 + z2` the dispersion model. Omitting
the second part, `y ~ x1 + x2`, gives constant dispersion.

## 3. A worked example: reading skills

We use the `ReadingSkills` data from the **betareg** package: the response
`accuracy` is a reading-accuracy score in $(0, 1)$ for 44 children, `dyslexia`
indicates a diagnosis of dyslexia, and `iq` is a standardised nonverbal IQ score.

```{r data}
if (has_betareg) {
  data("ReadingSkills", package = "betareg")
} else {
  # Synthetic stand-in so the vignette renders without 'betareg'.
  set.seed(1)
  n <- 44
  ReadingSkills <- data.frame(dyslexia = factor(rep(c("no", "yes"), each = n / 2)),
                              iq = rnorm(n))
  mu <- simplex_linkinv(1.0 - 0.9 * (ReadingSkills$dyslexia == "yes") +
                          0.4 * ReadingSkills$iq, "logit")
  ReadingSkills$accuracy <- rsimplex(n, mu, exp(0.5))
}
head(ReadingSkills)
```

We model the mean accuracy through `dyslexia` and `iq`, and let the **dispersion**
depend on dyslexia status — a genuine variable-dispersion model:

```{r fit}
fit <- fastsimplexreg(accuracy ~ dyslexia + iq | dyslexia,
                      data = ReadingSkills, link = "logit")
summary(fit)
```

Accuracy is lower for dyslexic children and increases with IQ; the dispersion
also differs by group. The fitted mean and dispersion are available directly:

```{r extract}
head(cbind(mu = fitted(fit), phi = fitted(fit, "dispersion")))
confint(fit)
```

### Diagnostics

The `plot()` method returns **ggplot2** panels (residuals vs fitted, a normal
Q-Q plot, scale-location, and observed vs fitted):

```{r diagnostics, fig.width = 7, fig.height = 5.5, out.width = "100%"}
plot(fit, which = 1:4)
```

### Choosing a link

Because all four mean links are supported, they can be compared by AIC:

```{r links}
links <- c("logit", "probit", "cloglog", "neglog")
aic <- sapply(links, function(lk)
  AIC(fastsimplexreg(accuracy ~ dyslexia + iq | dyslexia,
                     data = ReadingSkills, link = lk)))
round(sort(aic), 2)
```

## 4. Density and simulation

The distribution utilities are vectorised and share the C++ backend:

```{r dist-utils}
dsimplex(c(0.2, 0.5, 0.8), mu = 0.5, phi = 1)
set.seed(42)
y <- rsimplex(1e4, mu = 0.35, phi = 0.8)
c(mean = mean(y), target_mu = 0.35)   # sample mean approximates mu
```

## 5. Performance notes

The critical path is entirely in C++: the log-likelihood and its analytic score
(so no numerical differentiation during optimisation), a native BFGS that avoids
repeated R/C++ crossings, BLAS matrix-vector products for the linear predictors,
and an optional OpenMP-parallelised observation loop (`n_threads = 0` uses all
cores). Inference (the Hessian and standard errors) is computed only when
`inference = TRUE`, so exploratory fits on massive data can skip it.

## References

Barndorff-Nielsen, O. E. and Jørgensen, B. (1991). Some parametric models on the
simplex. *Journal of Multivariate Analysis*, **39**(1), 106-116.

Song, P. X.-K. and Tan, M. (2000). Marginal models for longitudinal continuous
proportional data. *Biometrics*, **56**(2), 496-502.

Zhang, P., Qiu, Z. and Shi, C. (2016). simplexreg: An R package for regression
analysis of proportional data using the simplex distribution. *Journal of
Statistical Software*, **71**(11), 1-21.
