---
title: "Simplex mixed models for nested proportion data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simplex mixed models for nested proportion data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(fastsimplexreg)
has_betareg <- requireNamespace("betareg", quietly = TRUE)
```

## 1. Nested proportion data

Continuous proportions in $(0, 1)$ are frequently **nested**: repeated
measurements within a subject, students within schools, plots within sites. The
observations within a cluster are correlated, and treating them as independent
understates uncertainty and can bias both the mean and the dispersion estimates.
`fastsimplexregmixed()` extends the simplex regression of
[`fastsimplexreg()`](fastsimplexreg.html) with a cluster-specific random effect,
while keeping the variable-dispersion structure.

## 2. The two-level simplex mixed model

For observation $i$ in cluster $j$, conditionally on a cluster random effect
$b_j \sim N_q(0, \Sigma)$,
$$
y_{ij} \mid b_j \sim \mathrm{Simplex}(\mu_{ij}, \phi_{ij}), \qquad
g(\mu_{ij}) = x_{ij}^\top\beta + z_{ij}^\top b_j, \qquad
\log\phi_{ij} = w_{ij}^\top\gamma .
$$
The fixed effects $\beta$ (mean) and $\gamma$ (dispersion) use the same
multi-part `Formula` as the fixed-effects fit; the $q$-dimensional random effect
$b_j$ enters the mean submodel and induces the within-cluster correlation. The
mean link $g$ is one of `logit`, `probit`, `cloglog`, `neglog`; the dispersion
uses a log link and carries fixed effects only.

Because the random effects are unobserved, the likelihood is the **marginal**
likelihood obtained by integrating them out,
$$
\ell(\theta) = \sum_j \log \int_{\mathbb{R}^q}
  \exp\!\Big\{\textstyle\sum_i \log f(y_{ij}\mid b)\Big\}\,
  N(b; 0, \Sigma)\, \mathrm{d}b ,
$$
which has no closed form.

### Adaptive Gauss-Hermite quadrature

The integral is approximated by **adaptive Gauss-Hermite quadrature** (AGHQ). For
each cluster the integrand is centred at its posterior mode and rescaled by the
curvature there, after which a Gauss-Hermite rule with `nAGQ` points per
dimension is applied. `nAGQ = 1` recovers the **Laplace approximation**; larger
`nAGQ` refines the integral, at the cost of `nAGQ^q` evaluations per cluster.
The random-effect covariance $\Sigma = D D^\top$ is estimated on an
unconstrained log-Cholesky scale, so the estimate is always positive definite.
The per-cluster inner mode-finding, the quadrature and the analytic score run in
C++ and are parallelised over clusters with OpenMP.

### The `random` interface

Random effects and the grouping factor are supplied through `random`, using the
lme4-style bar:

* `random = ~ 1 | subject` — a random intercept;
* `random = ~ 1 + t | subject` — a random intercept and a random slope in `t`;
* `random = ~ 0 + t | subject` — a random slope only.

## 3. A worked example: gasoline yield by crude-oil batch

We use the `GasolineYield` data from the **betareg** package: the response
`yield` is the proportion of crude oil converted to gasoline, measured across
experimental settings of temperature `temp` for 10 crude-oil `batch`es (2-4
measurements each). Yields from the same batch share the crude oil's unmeasured
properties — a natural nested structure. We model the mean yield through `temp`
and add a **random intercept per batch**.

```{r data}
if (has_betareg) {
  data("GasolineYield", package = "betareg")
} else {
  # Synthetic stand-in so the vignette renders without 'betareg'.
  set.seed(1)
  J <- 10; nj <- 3; n <- J * nj
  batch <- factor(rep(seq_len(J), each = nj))
  temp <- rep(c(205, 275, 345), J) + rnorm(n, 0, 5)
  b <- rnorm(J, 0, 0.5)[batch]
  mu <- simplex_linkinv(-2.5 + 0.006 * temp + b, "logit")
  GasolineYield <- data.frame(yield = rsimplex(n, mu, exp(-1)), temp = temp,
                              batch = batch)
}
str(GasolineYield)
```

```{r fit}
fit <- fastsimplexregmixed(
  yield ~ temp,
  random = ~ 1 | batch,
  data = GasolineYield,
  link = "logit",
  nAGQ = 15,
  n_threads = 1
)
summary(fit)
```

The estimated between-batch variance and the group-level random effects:

```{r varcorr}
VarCorr(fit)
head(ranef(fit))
```

A non-negligible random-intercept variance confirms that batches differ
systematically in baseline yield beyond what temperature explains — exactly the
heterogeneity a fixed-effects fit on the pooled data would ignore. The number of
groups and the marginal fit statistics:

```{r info}
c(groups = ngrps(fit), nobs = nobs(fit))
c(logLik = as.numeric(logLik(fit)), AIC = AIC(fit), BIC = BIC(fit))
```

### Conditional vs population predictions

By default `predict()` is **conditional** on the estimated random effects (a
batch-specific curve); `re.form = NA` gives the **population-level**
(marginal-mode) prediction with $b = 0$:

```{r predict}
b1 <- levels(GasolineYield$batch)[1]
newd <- data.frame(temp = c(250, 350, 450),
                   batch = factor(b1, levels = levels(GasolineYield$batch)))
data.frame(
  temp        = newd$temp,
  conditional = predict(fit, newdata = newd, type = "response"),
  population  = predict(fit, newdata = newd, type = "response", re.form = NA)
)
```

### Diagnostics

The `plot()` method mirrors the fixed-effects diagnostics, with residuals
computed **conditional on the empirical-Bayes random effects**:

```{r diagnostics, fig.width = 7, fig.height = 5.5, out.width = "100%"}
plot(fit, which = 1:4)
```

## 4. Choosing `nAGQ` and performance

`nAGQ` trades accuracy for cost. `nAGQ = 1` (Laplace) is fastest and is often
adequate when clusters are large; `nAGQ = 9`-`15` gives high accuracy for small
clusters. Since the tensor grid has $\mathtt{nAGQ}^q$ nodes, keep the number of
random effects small ($q \le 3$) and lower `nAGQ` as $q$ grows (good defaults:
`11` for $q = 1$, `9` for $q = 2$, `7` for $q = 3$). Because clusters are
conditionally independent, the work is parallelised over clusters — set
`n_threads = 0` to use all available cores.

## 5. Scope and extensions

This version supports a single grouping factor (two-level nesting), Gaussian
random effects in the mean submodel, and fixed-effect (variable) dispersion.
Crossed or three-level random effects, and random effects in the dispersion
submodel, are natural extensions for future releases.

## References

Barndorff-Nielsen, O. E. and Jørgensen, B. (1991). Some parametric models on the
simplex. *Journal of Multivariate Analysis*, **39**(1), 106-116.

Pinheiro, J. C. and Bates, D. M. (1995). Approximations to the log-likelihood
function in the nonlinear mixed-effects model. *Journal of Computational and
Graphical Statistics*, **4**(1), 12-35.

Song, P. X.-K. and Tan, M. (2000). Marginal models for longitudinal continuous
proportional data. *Biometrics*, **56**(2), 496-502.
